---
title: "Linear Model and ANOVA by Hand (CO2 Example)"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
---

## Goal

This document explains what `Theory/lm_and_aov_by_hand.R` does.
The script manually recreates core outputs of `summary(lm(...))` and `summary(aov(...))` for a linear model with an interaction term.

The fitted model is:

$$
\text{uptake}_i = \beta_0 + \beta_1 \cdot \text{conc}_i + \beta_2 \cdot \text{Treatment}_i + \beta_3 \cdot (\text{conc}_i \cdot \text{Treatment}_i) + \epsilon_i
$$

with normally distributed errors.

## 1. Data Preparation

The script uses the built-in `CO2` dataset:

- `predictor1`: `conc` (numeric concentration)
- `predictor2`: `Treatment` converted to 0/1 (`0 = nonchilled`, `1 = chilled`)
- `response`: `uptake`

```{r}
predictor1 <- CO2[["conc"]]
temp <- CO2[["Treatment"]]
predictor2 <- numeric(length(temp))
predictor2[temp == "chilled"] <- 1
response <- CO2[["uptake"]]

n <- length(response)
df_residual <- n - 4  # 4 coefficients: intercept, conc, treatment, interaction
c(n = n, df_residual = df_residual)
```

## 2. Fit the Model by Maximizing Log-Likelihood

The script defines a log-likelihood under a Gaussian error model and uses `optim(..., method = "BFGS")` to estimate coefficients.

```{r}
ll <- function(bs) {
  sum(
    dnorm(
      response,
      mean = bs[1] + bs[2] * predictor1 + bs[3] * predictor2 + bs[4] * (predictor1 * predictor2),
      log = TRUE
    )
  )
}

op <- optim(
  par = c(1, 0.1, 0.1, 0.01),
  fn = ll,
  method = "BFGS",
  control = list(fnscale = -1, reltol = 1e-12),
  hessian = TRUE
)

Estimate <- op$par
Estimate
```

## 3. Residuals and Coefficient Inference

After obtaining coefficients, the script computes:

- fitted values and residuals
- variance-covariance matrix from the Hessian
- residual variance scaling (important because the Hessian corresponds to unit variance)
- standard errors, t-values, and p-values

```{r}
predicted <- Estimate[1] + Estimate[2] * predictor1 + Estimate[3] * predictor2 + Estimate[4] * (predictor1 * predictor2)
residuals <- response - predicted

var_covar_mat <- solve(-op$hessian)
sigma2_hat <- sum(residuals^2) / df_residual
var_covar_mat_scaled <- sigma2_hat * var_covar_mat
std_error <- sqrt(diag(var_covar_mat_scaled))

t_values <- Estimate / std_error
p_values <- 2 * pt(-abs(t_values), df = df_residual)

Coefficients <- do.call(cbind, list(Estimate, std_error, t_values, p_values)) |> as.data.frame()
names(Coefficients) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")
row.names(Coefficients) <- c("Intercept", "conc", "Treatment", "conc:Treatment")
knitr::kable(Coefficients, digits = 4, caption = "Manual coefficient table")
```

Interpretation:

- Large absolute t-values indicate stronger evidence against $H_0:\beta_j=0$.
- Small p-values indicate terms that contribute to explaining `uptake`.

## 4. Model Fit Metrics

The script then recreates key model-level diagnostics:

- residual standard error
- $R^2$ and adjusted $R^2$
- overall F-test for whether predictors improve fit over intercept-only model

```{r}
residual_standard_error <- sqrt(sum(residuals^2) / df_residual)

sum_sq_total <- sum((response - mean(response))^2)
sum_sq_res <- sum(residuals^2)
r_squared <- 1 - sum_sq_res / sum_sq_total
adj_r_squared <- 1 - (1 - r_squared) * ((n - 1) / df_residual)

df_model <- 3  # conc, Treatment, interaction
ms_model <- (sum_sq_total - sum_sq_res) / df_model
ms_res <- sum_sq_res / df_residual
f_statistic <- ms_model / ms_res
f_p_value <- pf(f_statistic, df_model, df_residual, lower.tail = FALSE)

fit_summary <- data.frame(
  Metric = c("Residual standard error", "R-squared", "Adjusted R-squared", "F-statistic", "F-test p-value"),
  Value = c(residual_standard_error, r_squared, adj_r_squared, f_statistic, f_p_value)
)
knitr::kable(fit_summary, digits = 5, caption = "Manual model-level statistics")
```

## 5. Cross-Check with `lm()` and `aov()`

Finally, the script compares manual calculations to built-in R model summaries.

```{r}
model <- lm(uptake ~ conc * Treatment, data = CO2)
summary(model)
summary(aov(model))
```

These outputs should align with the manually computed coefficients, uncertainty estimates, and overall F-test (up to small numerical differences).

## 6. ANOVA Logic Repeated Explicitly

The last block in the script repeats ANOVA quantities directly from sums of squares:

$$
SS_{Total} = SS_{Model} + SS_{Residual}
$$

with:

- $MS_{Model} = SS_{Model}/df_{Model}$
- $MS_{Residual} = SS_{Residual}/df_{Residual}$
- $F = MS_{Model}/MS_{Residual}$

This makes clear how ANOVA and linear regression summaries are connected for Gaussian linear models.
