---
title: "Randomization Workflow for Finite Groups"
format:
  pdf:
    number-sections: true
    callout-appearance: simple
fontsize: 11pt
geometry: margin=1in
---

## Motivation and overview

This document illustrates the **new finite-group randomization approach**
using a realistic experimental setup with multiple blocking factors and continuous covariates.
The goal is to assign a *finite set of animals*
(each with measured properties such as weight and blood values)
to a predefined experimental design (treatments × locations)
such that the resulting groups are as **balanced as possible** with respect to these covariates.
Unlike classical stratification-based randomization,
the approach shown here does **not discretize** continuous variables.
Instead, it performs a **global optimization over all possible assignments**,
directly operating on the continuous data.

\noindent\fbox{%
  \parbox{\textwidth}{%
  \textbf{Why is ast2ast important?}

  \medskip
  \textbf{ast2ast} is a key technical foundation for the sustainable development of
  OpenStats and related scientific software. It automatically translates selected R code
  into efficient compiled code, so users keep the familiar R interface while demanding
  computations run much faster.

  \medskip
  The benchmark below illustrates this practical benefit: for a simple assignment problem
  (48 experimental units allocated to three treatment groups using two covariates, weight and blood),
  the compiled implementation runs about \textbf{10$\times$ faster} than the reference R version.
  For larger and more complex designs the advantage is even more pronounced; however,
  those cases can take tens of minutes in pure R and are therefore not included here.

  \medskip
  \begin{center}
  \includegraphics[
    width=0.95\linewidth,
    height=0.55\textheight,
    keepaspectratio
  ]{benchmark.png}
  \end{center}
  }
}

\newpage

## Experimental design and available animals

We first define the **target design**.
Each row corresponds to a concrete experimental slot defined by its treatment and location.

```{r cache=TRUE}
n_blocks <- data.frame(
  treatment = rep(c("control", "one", "two"), each = 64L),
  location  = rep(rep(c("A", "B", "C", "D"), each = 16L), 3L)
)
knitr::kable(table(n_blocks))
```

Next, we generate the available animals together with their measured covariates.
Normally, these properties would be measured.

```{r cache=TRUE, results="hide"}
#| echo: false
set.seed(3344)
groups <- data.frame(
  weights = rnorm(192, 300, 25),
  blood   = rnorm(192, 100, 1)
)
temp <- stack(groups)
library(ggplot2)
ggplot(data = temp, aes(y = values)) +
  geom_boxplot() +
  facet_wrap(~ ind, scales = "free")
```

\newpage

## Finite-group randomization by global optimization

The core step is the call to `random_finite_assign()`.
Conceptually, this function searches over permutations of animals and assigns each
animal **exactly once** to one slot in the design, minimizing
imbalance across all covariates and block factors simultaneously.

```{r cache=TRUE, results="hide"}
res <- Randomization::random_finite_assign(
  seed = 1234,
  groups = groups,
  design = n_blocks,
  max_iter = 1000L
)
```

## Numerical assessment of balance

As a first sanity check, we summarize means and standard deviations
of the covariates by treatment and by location.

```{r}
#| echo: false
df <- res$assigned
df$treatment <- n_blocks$treatment
df$location  <- n_blocks$location
t <- aggregate(cbind(weights, blood) ~ treatment, df,
               \(x) c(mean = mean(x), sd = sd(x)))
l <- aggregate(cbind(weights, blood) ~ location, df,
               \(x) c(mean = mean(x), sd = sd(x)))

knitr::kable(as.matrix(t))
knitr::kable(as.matrix(l))
```

\newpage

## Visual assessment using joint distributions

Because balancing marginal means alone can be misleading,
we also inspect the **joint distribution** of `weights` and `blood`.

```{r}
#| echo: false
library(ggplot2)

ggplot(df, aes(x = weights, y = blood)) +
  geom_density_2d() +
  facet_grid(location ~ treatment)

ggplot(df, aes(x = weights, y = blood)) +
  geom_point(alpha = 0.3) +
  geom_density_2d() +
  facet_wrap(~ treatment)

ggplot(df, aes(x = weights, y = blood)) +
  geom_point(alpha = 0.3) +
  geom_density_2d() +
  facet_wrap(~ location)
```

\newpage

## How this differs from the previous stratification-based approach

The previous workflow relied on:
- discretizing continuous covariates into quantiles,
- defining strata based on these bins,
- and randomizing within each stratum.

While effective for simple cases, this approach becomes problematic when:
- No large surplus of experimental units relative to the number of experiments slots
- multiple continuous covariates are involved,
- interactions between blocking factors matter,
- or the design is unbalanced.

The **finite-group approach shown here differs fundamentally**:
- No discretization of covariates is required.
- Balance is evaluated directly on the continuous scale.
- The assignment is optimized **globally**, not stratum by stratum.
- Interactions between factors (e.g. treatment × location) are handled naturally.

In short, instead of asking
*“How do we randomize within bins?”*,
we ask *“Given this exact design and these exact animals, what is the best possible assignment?”*
