---
title: "Power Analysis and Randomization Workflow"
format:
  pdf:
    number-sections: true
fontsize: 11pt
geometry: margin=1in
---


# Introduction

This document demonstrates an end-to-end workflow for planning and executing
biological experiments using the R package *Randomization*.
It covers three core components:

1. **Power analysis**
   We determine how many repetitions per treatment level are required to reach
   a desired statistical power (here: one-way ANOVA based on *Cohen’s f*).
2. **Design matrix generation**
   From user-defined predictors (e.g., Treatment, Cell line), a full design
   matrix is created. All experimental combinations are expanded to the
   repetition count determined by the power analysis.
3. **Random assignment**
   The design matrix is randomly assigned to user-defined groups
   (e.g., measurement days, plates, runs).
   The following methods are implemented:
   - **Simple randomisation**
     Every row is independently assigned according to the specified ratio.
   - **Block randomisation**
     Randomisation occurs *within* levels of a chosen factor
     (e.g., Treatment-wise balancing across days).
   - **Block-stratified randomisation**
     Randomisation occurs within *combinations* of several predictors  
     (e.g., stratifying by Treatment × Cell line).
   - **Balanced randomisation**
     Each predictor combination is balanced across groups, followed by a global shuffle.

```{r}
#| echo: false
sr <- function(res) {
  res <- res[, -3] # remove id
  rt <- table(res)
  lapply(c("Day1", "Day2", "Day3"), function(d) {
    sub <- rt[, , d]
    rs <- rowSums(sub)
    cs <- colSums(sub)
    sub <- rbind(sub, Sum = cs)
    sub <- cbind(sub, Sum = c(rs, sum(cs)))
    cat(d, "\n")
    print(sub)
  })
}
```

{{< pagebreak >}}

# Power analysis & Random assignment

In this chapter, the usage of function `randomization_one_way_anova` is demonstrated. The results are printed by an internal helper function called *sr*.

## Predictors

First the Predictors have to be defined as a named list.
Each element represents one experimental factor and its possible levels.
The *primary_factor* determines which predictor is used in the one-way ANOVA
power analysis. Only this factor influences the required sample size.
All other predictors are later included only in the design matrix.

```{r}
predictors <- list(
  cellLines = c("HeLa", "Hek"),
  Treatment = LETTERS[1:4]
)
primary_factor <- "Treatment"
```

## Simple randomisation

Each row of the design matrix is randomly assigned to one of the groups.
Assignments follow the user-defined ratios (e.g., 1:1:1).
There is no balancing by Treatment or Cell line — this is the most basic form
of randomisation.

```{r}
res <- Randomization::randomization_one_way_anova(
  predictors, "Treatment", groups = c("Day1", "Day2", "Day3"),
  group_type = "infinite",
  ratios = c(1, 1, 1), 0.25, 0.05, 0.8,
  randomization_method = "simple", seed = 42
) |> sr()
```

{{< pagebreak >}}

## Block randomisation

Here, randomisation is performed *within each level* of the primary factor.
For example:
- each Treatment level is randomised separately,
- ensuring that every Treatment is distributed across measurement days
  according to the target ratios.
This reduces confounding between Treatment and Day.

```{r}
res <- Randomization::randomization_one_way_anova(
  predictors, "Treatment", groups = c("Day1", "Day2", "Day3"),
  group_type = "infinite",
  ratios = c(1, 1, 1), 0.25, 0.05, 0.8,
  randomization_method = "block", seed = 42
) |> sr()
```

{{< pagebreak >}}

## Block-stratified randomisation

This method stratifies on *multiple predictors simultaneously* (e.g., Treatment × Cell line).
Randomisation happens independently within every stratum.
This is extremely useful when laboratories want to ensure that
all combinations (such as all Treatment–Cell line pairs)
appear on all days in a controlled ratio.

```{r}
res <- Randomization::randomization_one_way_anova(
  predictors, primary_factor = "Treatment",
  groups = c("Day1", "Day2", "Day3"),
  group_type = "infinite", ratios = c(1, 1, 1),
  cohens_f = 0.25, sig_level = 0.05, desired_power = 0.8,
  randomization_method = "block_stratified",
  strata_cols = c("cellLines", "Treatment"), seed = 42
) |> sr()
```

{{< pagebreak >}}

# Advanced Randomized Assignment

The following section illustrates the full analytical workflow step by step.

1. We compute the required sample size using `pwr.anova.test()` from the *pwr* package.
2. We construct a completely randomised design matrix based on all predictors.
3. We optionally expand this matrix to meet the sample size requirements.
4. We apply a chosen randomisation method to assign each row to a group.

## Power Analysis

```{r}
n_levels <- length(predictors[[primary_factor]])
cohens_f <- 0.25
sig_level <- 0.05
desired_power <- 0.8
power_analysis_res <- pwr::pwr.anova.test(
  k = n_levels,
  f = cohens_f,
  sig.level = sig_level,
  power = desired_power
)
plot(power_analysis_res)

```

## Completely randomized design

A fully crossed design matrix is generated using `expand.grid()`.
This includes every combination of predictors, e.g.:

- 2 cell lines
- 4 treatments
→ 8 combinations in total

The required number of repetitions per treatment level is calculated from the
power analysis. These repetitions are then distributed across predictor
combinations, producing the final design matrix used in randomisation.

```{r}
n_per_level <- power_analysis_res$n
knitr::kable(data.frame("N per level" = n_per_level))
other_factors <- predictors[setdiff(names(predictors), primary_factor)]
combos_per_level <- prod(vapply(other_factors, length, integer(1)))
n_per_combo <- ceiling(n_per_level / combos_per_level)
design <- Randomization:::completely_randomised_design(predictors, n_per_combo)
design_excerpt <- table(design) |> as.data.frame()
design_excerpt <- design_excerpt[design_excerpt$Freq != 0, ]
knitr::kable(design_excerpt[, -3])
```

{{< pagebreak >}}

## Random group assignment

The design matrix is assigned to groups using the unified `random_assign()` function.
The example below demonstrates block-stratified randomisation by stratifying on `cellLines`,
i.e. the assignments within each cell line follow the target group ratios.

```{r}
groups <- c("Day1", "Day2", "Day3")
ratios <- c(1, 1, 1)
col <- "Group"
strata_cols <- "cellLines"
design_assigned <- Randomization::random_assign(
  df = design, groups = groups, group_type = "infinite", ratios = ratios,
  col = col, strata_cols = strata_cols,
  randomization_method = "block_stratified", seed = 1234
) |> sr()
```
