
---
title: "Randomization Strategies in Experimental Design"
author: "Konrad Krämer"
format:
  html:
    toc: true
    number-sections: true
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This document demonstrates currently implemented randomization approaches in the `Randomization` package:

1. Generation of a completely randomized factorial design  
2. Finite assignment using loss minimization  
3. Block-stratified randomization  
4. Diagnostics via graphical inspection  

We further discuss the theoretical foundation of the implemented loss functions.

---

# Experimental Design Matrix

We consider a 2 × 3 factorial design:

- Treatments: control vs ibo
- Locations: Mannheim, Karlsruhe, Aachen
- 16 animals per treatment × location combination

```{r}
#| echo: false
library(Randomization)
set.seed(1234)
treatments <- c("control", "Compound1", "Compound2")
locations <- c("Mannheim", "Karlsruhe", "Aachen")
n <- 16L
n_per_location <- n * length(treatments)
df <- Randomization:::completely_randomised_design(
  predictors = list(treatments = treatments, locations = locations), n_per_level = n
)
knitr::kable(table(df$treatments, df$locations))
dfs <- split(df, df$locations)
df <- dfs[[3L]]
```

---

# Finite Assignment via Loss Minimization

We generate animals with two covariates:

- Weight
- Blood value

```{r}
set.seed(1234)
groups <- data.frame(
  weights = rnorm(n_per_location, 300, 40),
  blood_value = rnorm(n_per_location, 10, 2.5)
)
result <- Randomization::random_finite_assign(seed = 1234, groups = groups, design = df, verbose = FALSE)
df$weights <- result$assigned$weights
df$blood_value <- result$assigned$blood_value
df$animal_ids <- result$assigned$unit_index
```

## Diagnostic Plots

```{r}
#| echo: false
library(ggplot2)
l <- length(groups$weights)
d <- data.frame(
  treatments = rep(df$treatments, 2L),
  weights = c(df$weights, groups$weights),
  groups = c(rep("Optimized", l), rep("Original", l))
)
p1 <- ggplot(data = d,
  aes(x = treatments, y = weights, shape = groups, fill = groups, groups = groups)) +
  geom_boxplot() +
  stat_summary(
    fun = mean, geom = "point", position = position_dodge(width = 0.8),
    size = 2, color = "black"
  ) +
  stat_summary(
    fun = mean, geom = "line", aes(group = groups, linetype = groups),
    position = position_dodge(width = 0.8), color = "black"
  )
d <- data.frame(
  treatments = rep(df$treatments, 2L),
  blood_value = c(df$blood_value, groups$blood_value),
  groups = c(rep("Optimized", l), rep("Original", l))
)
p1
p2 <- ggplot(data = d,
  aes(x = treatments, y = blood_value,
      shape = groups, fill = groups, groups = groups)) +
  geom_boxplot() +
  stat_summary(
    fun = mean, geom = "point", position = position_dodge(width = 0.8),
    size = 2, color = "black"
  ) +
  stat_summary(
    data = d, aes(group = groups, linetype = groups),
    fun = mean, geom = "line",
    position = position_dodge(width = 0.8), color = "black"
  )
p2
```

---

# Block-Stratified Randomization

We now assign experimental blocks (8 blocks per location).

```{r}
experimental_blocks <- paste0("block_", 1:8)
ratios <- rep(1, 8)

df <- Randomization::random_assign(
  df = df, groups = experimental_blocks, ratios = ratios,
  col = "block", strata_cols = c("treatments", "locations"),
  randomization_method = "block_stratified", seed = 1234
)
```

## Block Diagnostics

```{r}
#| echo: false
res <- as.data.frame.matrix(table(df$treatments, df$block))
knitr::kable(res)
```

---

# Theoretical Foundation of Loss Functions

{{< include moments_loss_function.qmd >}}

---



{{< include mahalanobis_distance_loss_function.qmd >}}


