---
title: "Randomization Workflow for finite groups"
format:
  pdf:
    number-sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r}
#| echo: false
library(ggplot2)

pl <- function(res) {
  res$random_group <- as.numeric(res$random_group)
  global_df <- data.frame(
    value = weights,
    source = "Global"
  )
  block_df <- data.frame(
    value = res$random_group,
    source = "Assigned",
    Treatment = res$Treatment,
    cellLines = res$cellLines
  )
  p1 <- ggplot() +
    geom_density(data = global_df,
      aes(x = value, linetype = source),
      linewidth = 1) +
    geom_density(data = block_df,
      aes(x = value, color = Treatment),
      alpha = 0.6) +
    labs(
      title = "Global vs Block-wise Weight Distributions",
      x = "Weight",
      y = "Density"
    ) +
    theme_minimal()
  p2 <- ggplot() +
    geom_density(data = global_df,
      aes(x = value, linetype = source),
      linewidth = 1) +
    geom_density(data = block_df,
      aes(x = value, color = cellLines),
      alpha = 0.6) +
    labs(
      title = "Global vs Block-wise Weight Distributions",
      x = "Weight",
      y = "Density"
    ) +
    theme_minimal()
  cowplot::plot_grid(p1, p2, ncol = 1L)
}
```

# Introduction

This document demonstrates an end-to-end workflow for planning and executing
biological experiments using the R package *Randomization*.
It covers three core components:

1. **Power analysis**
   We determine how many repetitions per treatment level are required to reach
   a desired statistical power (here: one-way ANOVA based on *Cohen’s f*).
2. **Design matrix generation**
   From user-defined predictors (e.g., Treatment, Cell line), a full design
   matrix is created. All experimental combinations are expanded to the
   repetition count determined by the power analysis.
3. **Random assignment**
   The design matrix is randomly assigned to user-defined groups
   (e.g., measurement weights, plates, runs).
   In this document the assignment of a finite group (*weights*) is demonstrated.

```{r}
predictors <- list(
  cellLines = c("HeLa", "Hek"),
  Treatment = LETTERS[1:4]
)
primary_factor <- "Treatment"
set.seed(1234)
weights <- rnorm(250, 300, 25)
```

{{< pagebreak >}}

## Simple randomisation

Each row of the design matrix is randomly assigned to one of the groups.
There is no balancing by Treatment or Cell line — this is the most basic form
of randomisation. Here the weights of e.g. animals are assigned to the design matrix.

```{r}
Randomization::randomization_one_way_anova(
  predictors = predictors, primary_factor = "Treatment",
  groups = weights, group_type = "finite",
  cohens_f = 0.25, sig_level = 0.05, desired_power = 0.9,
  randomization_method = "simple", n_quantiles = 11L,
  seed = 42
) |> pl()
```

{{< pagebreak >}}

## Block randomisation

Here, randomisation is performed *within each level* of the primary factor.
For example:
- each Treatment level is randomised separately,
- the weights of e.g. animals are assigned to the design matrix.

```{r}
Randomization::randomization_one_way_anova(
  predictors = predictors, primary_factor = "Treatment",
  groups = weights, group_type = "finite",
  cohens_f = 0.25, sig_level = 0.05, desired_power = 0.9,
  randomization_method = "block", n_quantiles = 11L,
  seed = 124
) |> pl()
```

{{< pagebreak >}}

## Block-stratified randomisation

This method stratifies on *multiple predictors simultaneously* (e.g., Treatment × Cell line).
Randomisation happens independently within every stratum.
Here the weights of e.g. animals are assigned to the design matrix.

```{r}
Randomization::randomization_one_way_anova(
  predictors = predictors, primary_factor = "Treatment",
  groups = weights, group_type = "finite",
  cohens_f = 0.25, sig_level = 0.05, desired_power = 0.9,
  randomization_method = "block_stratified", n_quantiles = 11L,
  strata_cols = c("cellLines", "Treatment"),
  seed = 123
) |> pl()
```
